### 2021-04-19 submission/manuscript.Rmd
Update results description for 5-day PEG PERMANOVA

``` {r 5-day PEG microbiota}
f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
f_d_top_2_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
  top_n(2, R2) %>% #Select top 2 variables based on R2
  select(effects, R2, `Pr(>F)`) #Select relevant column names

f_d_top_contrib= (f_d_top_2_contrib[1,1])
f_d_sec_contrib = (f_d_top_2_contrib[2,1])
f_d_top_R2_sum = f_d_top_2_contrib %>% 
  select(R2) %>% sum
f_d_top_p <- f_d_top_2_contrib %>% 
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text
f_d_top_R2 = f_d_top_2_contrib[1,2]
f_d_sec_R2 = f_d_top_2_contrib[2,2]

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A) with the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B). Next, we examined  alpha diversity by looking at Shannon diversity index over time, we see that although both clindamycin and PEG treatments decrease diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by analyzing what changed in mice treated with PEG alone between baseline and one day post-treatment. We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the impact on *Bacteroides* was different compared to clindamycin treated mice (Fig. 2D). Thus, PEG has detrimental impacts on the microbiome that are maintained over time.
Surprisingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine what other bacteria were changing during the post-infection period when *C. difficile* changed from undectable to detectable. Interestingly, we found ___ bacteria were shifting during the time period when C. difficile was becoming detectable in the stools (Fig. S1)

@@

\includegraphics{figure_2.pdf}
**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. 
\newpage

>>>>

``` {r 5-day PEG microbiota}
f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
f_d_sign_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
  filter(`Pr(>F)` < 0.05) #Only select significant contributors

#List of variables that significantly explained variation in stool samples
f_d_top_effects <- f_d_sign_contrib %>% 
  pull(effects)
#Sum of R2 values for variables that significantly explained variation in stool samples
f_d_top_R2_sum <- f_d_sign_contrib %>% 
  select(R2) %>% sum
#P-value for variables that significantly explained variation in stool samples
f_d_top_p <- f_d_sign_contrib %>% 
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B). Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria that differ across treatment groups over multiple timepoints. We found... (Fig. 2E). Thus, PEG has a significant impact on the fecal microbiota that was maintained over time.

Surprisingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine what other bacteria were changing during the post-infection period when *C. difficile* changed from undectable to detectable. Interestingly, we found ___ bacteria were shifting during the time period when C. difficile was becoming detectable in the stools (Fig. S1)

@@

\includegraphics{figure_2.pdf}
**Figure 2. 5-day PEG treatment disrupts the stool microbiota for a longer amount of time compared to clindamycin-treated mice.** A. PCoA of Bray-Curtis distances from stool samples collected throughout the experiment.
\newpage

### 2021-04-19 submission/manuscript.tex
Update results description for 5-day PEG PERMANOVA

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities (PERMANOVA combined R\textsuperscript{2} =
0.50, \emph{P} \textless{} 0.001) with the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (Data Set S1,
Sheet X). Cage effects refer to the well-documented phenomenon that mice
housed in the same cages have similar microbial communities due to
copraphagy, we tried to minimize the impact of cage effects on our
experiment by breaking up cagemates when assigning mice to treatment
groups and primarily housing only two mice per cage. Importantly,
although we conducted a total of 5 separate experiments, the experiment
number and it's interaction with treatment group was not one of the
variables that significantly explained the observed variation in fecal
communities (Data Set S1, Sheet X). When we examine alpha diversity by
looking at Shannon diversity index over time, we see that although both
clindamycin and PEG treatments decrease diversity, the diversity index
is less in the groups of mice that received PEG treatment compared to
those that received clindamycin through thirty days post-infection.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given previous work has shown
that PEG treatment disrupts the mucus layer, we decided to examine the
impact of the laxative on the mucosal microbiota and infection severity
by evaluating histopathology.

\begin{itemize}
\item
  Figure 2. 5-day PEG treatment disrupts the stool microbiota for a
  longer amount of time compared to clindamycin-treated mice.
\item
  Figure S1. 5-day PEG treatment plus 10-day recovery mice microbiota
  dynamics post-infection.
\item
  Figure 3. 5-day PEG treatment does not result in more severe CDIs,
  although mucosal microbiota is altered.
\item
  Figure 4. 1-day PEG treatment renders mice susceptible to transient
  \emph{C. difficile} colonization.
\item
  Figure 5. 1-day PEG treatment post \emph{C. difficile} challenge
  prolongs colonization regardless of whether an FMT is also
  administered.
\item
  Figure 6. Specific microbiota features associated with prolonged
  \emph{C. difficile} colonization in PEG treated mice.
\item
  Figure S2. Specific OTUs associated with clearance that are mostly
  absent in mice with prolonged \emph{C. difficile} colonization.

  \begin{itemize}
  \tightlist
  \item
    Ex. \emph{Muribaculum intestinale}.
  \end{itemize}
\item
  Figure 7. Schematic summarizing findings.
\end{itemize}

@@

\begin{itemize}
\item
  Summary of major findings
\item
  Discussion of prolonged persistence. C. difficile sequences detected
  in tissue samples. Association with mucin-degrading bacteria suggested
  by recent papers.
\item
  Discuss why we might not have observed more severe histology in PEG
  mice relative to clindamycin-treated mice

  \begin{itemize}
  \tightlist
  \item
    Antibiotics may also impact mucus layer
  \item
    Strain of bacteria used
  \end{itemize}
\item
  Protective bacteria missing in PEG-treated mice
\item
  Discuss what these findings might mean for human patients

@@

\includegraphics{figure_2.pdf} \textbf{Figure 2. 5-day PEG treatment
disrupts the stool microbiota for a longer amount of time compared to
clindamycin-treated mice.} A. \newpage

@@

\includegraphics{figure_5.pdf} \newpage
\includegraphics{figure_5_16S.pdf} \textbf{Figure 5. 1-day PEG treatment
post C. difficile challenge prolongs colonization regardless of whether
an FMT is also administered.} A. \newpage

>>>>

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities with most of the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (PERMANOVA
combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
phenomenon that mice housed in the same cages have similar microbial
communities due to copraphagy, we tried to minimize the impact of cage
effects on our experiment by breaking up cagemates when assigning mice
to treatment groups and primarily housing only two mice per cage.
Importantly, although we conducted a total of 5 separate experiments,
the experiment number and it's interaction with treatment group was not
one of the variables that significantly explained the observed variation
in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
treatment groups recovered to their baseline community structure either
10 or 30 days post-infection suggesting other community features besides
recovery to baseline were responsible for the prolonged \emph{C.
difficile} colonization in PEG-treated mice (Fig. 2B). Next, we examined
alpha diversity by looking at Shannon diversity index over time,
although both clindamycin and PEG treatments decreased diversity, the
diversity index was lower in the groups of mice that received PEG
treatment compared to those that received clindamycin through thirty
days post-infection (Fig. 2C). We next examined the bacterial genera
that shifted after PEG treatment by comparing the baseline samples of
mice treated with only PEG to samples from the same mice one day
post-treatment. We found 18 OTUs that were altered by PEG, the majority
of these OTUs decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment as Bacteroides actually decreased in clindamycin
treated mice (Fig. 2D). Finally, we examined the bacteria that differ
across treatment groups over multiple timepoints. We found\ldots{} (Fig.
2E). Thus, PEG has a significant impact on the fecal microbiota that was
maintained over time.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine what other bacteria were
changing during the post-infection period when \emph{C. difficile}
changed from undectable to detectable. Interestingly, we found \_\_\_
bacteria were shifting during the time period when C. difficile was
becoming detectable in the stools (Fig. S1)

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given that previous work had
demonstrated that PEG treatment disrupts the mucus layer and alters the
immune response in mice (16), we decided to examine the impact of PEG
treatment on the mucosal microbiota and CDI severity. We examined CDI
severity by evaluating cecum and colon histopathology {[}{]}.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (ref),
we also sequenced the tissue and stools of mock-challenged mice.
Examining the stools of the mock-infected PEG- and clindamycin-treated
mice revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S2A). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig. S2B).
Thus, most of the microbiota alterations we observed in the PEG-treated
mice were a result of the laxative and not an interaction between the
laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption} + Explain
motivation for experiment, set-up of experiment (Fig. 4A). + Transient
\emph{C. difficile} colonization in 1-day PEG treated mice with all PEG
mice clearing by 7dpi (Fig. 4B) + PCoA results (Day has larger R2 than
treatment group). PEG mice are closer to baseline communites after 7-day
period (Fig. 4C) + Shannon results show alpha diversity is only
transiently disrupted (Fig. 4D) + Highlight bacteria that are disrupted
by PEG or clindamycin treatment but recover within 7 days (Fig. 4E)

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered} + Reiterate motivation for experiment. Saw only a
transient microbiota disruption with 1-day PEG treatment alone. Also,
see if adding FMT helps with clearance, mention groups have been using
PEG or Abx/PEG treatments as a tool to engraft mice with human bacteria
communities. Set up of experiment (Fig. 5A). + Prolonged \emph{C.
difficile} colonization in post-CDI PEG treated mice (Fig. 5B) + FMT
appears to partially restore alpha diversity (Shannon, but not richness
Fig. 5C-D) + PCoA (necessary?) Could comment on the clustering of
Clindamycin and PEG-treated mice that received FMT (Fig. 5E) + Only 2
genera significantly impacted by FMT treatment, likely not as important
for clearance (Fig. 5F) + Bacteria that consistently differ between
groups over time, associated with prolonged colonization (Fig. G)

\textbf{Five-day post-infection community data can predict mice that
will have prolonged \emph{C. difficile} colonization} After identifying

\begin{itemize}
\item
  Figure 6. Specific microbiota features associated with prolonged
  \emph{C. difficile} colonization in PEG treated mice.
\item
  Figure S2. Specific OTUs associated with clearance that are mostly
  absent in mice with prolonged \emph{C. difficile} colonization.

  \begin{itemize}
  \tightlist
  \item
    Ex. \emph{Muribaculum intestinale}.
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Summary of major findings (Fig. 7A)
\item
  Discussion of prolonged persistence. C. difficile sequences detected
  in tissue samples. Association with mucin-degrading bacteria suggested
  by recent papers.
\item
  Discuss why we might not have observed more severe histology in PEG
  mice relative to clindamycin-treated mice

  \begin{itemize}
  \tightlist
  \item
    Antibiotics may also impact mucus layer
  \item
    Strain of bacteria used
  \end{itemize}
\item
  Protective bacteria missing in PEG-treated mice
\item
  Discuss what these findings might mean for human patients (Fig. 7B)

@@

\includegraphics{figure_2.pdf} \textbf{Figure 2. 5-day PEG treatment
disrupts the stool microbiota for a longer amount of time compared to
clindamycin-treated mice.} A. PCoA of Bray-Curtis distances from stool
samples collected throughout the experiment. \newpage

@@

\includegraphics{figure_5.pdf} A. Experimental timeline for groups
receiving PEG treatment post clindamycin gavage. B. Median and
individual CFU/g of C. difficile measured over time via culture.
Asterisks indicate time points with significant differences between
groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
Limit of detection is 1e2 CFU/g. Background shading indicates PEG
treatment period for applicable groups. Opacity of points reflects time
point. C. Median and individual Shannon Diversity measured over time.
Asterisks, background shading, and opacity of points consistent with
(B). D. Median and individual richness measured over time. Asterisks,
background shading, and opacity of points consistent with (B).

\includegraphics{figure_5_16S.pdf} \textbf{Figure 5. 1-day PEG treatment
post C. difficile challenge prolongs colonization regardless of whether
an FMT is also administered.} A. PCoA of all groups + FMT plotted over
time. Opacity of points reflects time point. B. Relative abundance of
genera significantly different over multiple time points post FMT or PBS
treatment, plotted over time. Limit of detection is .1\%. C. Relative
abundances of top 6 significant genera ranked by number of days
significant, plotted over time. Limit of detection is .1\%.
